---
title: "Final Project Join Data: OSHA Injuries and UV Exposure Data in Assessing Liability"
format: 
  dashboard: 
    theme: minty
    embed-resources: true
    fontsize: 16px
editor: 
  markdown: 
    wrap: sentence
---

```{r}
#| echo: false
#| warning: false
#| message: false

library(tidyverse) 
library(readxl)
library(janitor)
library(ggplot2)
library(maps)
library(plotly)

# Color Palette

uv_colors <- c(
  "Low" = "#2E7D9A",       
  "Medium" = "#E8A537",    
  "High" = "#C1495B"      
)

#loading osha_df 
osha_url <- "https://docs.google.com/spreadsheets/d/1GM90yTIXVE7xZoUrLwrqAO8yVWJsZtjNnkGbvDZhGpQ/export?format=csv&gid=1573088179"
osha_df <- read.csv(osha_url)

## Checking to make OSHA state col values in uppercase
osha_df <- osha_df %>%
  mutate(state = toupper(state))

#Loading NIH Exposure uv_df 
uv_df <- read_excel("uv-county.xlsx") %>%
    clean_names()

# Converting "statename" col in uv_df to abbreviations and matching with osha_df
uv_df <- uv_df %>%
  mutate(state = state.abb[match(statename, state.name)])

# Grouping by State-Level UV Exposure 
uv_state <- uv_df %>%
  group_by(state) %>%
  summarize(mean_uv = mean(uv_wh_m2_1961_1990, na.rm = TRUE))

#Joining both datasets
joined_df <- osha_df %>%
  left_join(uv_state, by = "state")

# Created a UV mean category in joined_df that classifies UV intensity in the state by Low, Medium or High 
joined_df <- joined_df %>%
    mutate(uv_category = case_when(
        mean_uv < quantile(mean_uv, 0.33, na.rm = TRUE) ~ "Low", 
        mean_uv < quantile(mean_uv, 0.66, na.rm = TRUE) ~ "Medium", 
        TRUE ~ "High"
    )) %>%
    mutate(uv_category = factor(uv_category, levels = c("Low", "Medium", "High"), ordered = TRUE))

# Created total skin disorders per state and avg uv variable
skin_uv_summary <- joined_df |>
    group_by(state) |>
    summarise(
        avg_uv = mean(mean_uv, na.rm = TRUE),
        skin_dis_count = sum(total_skin_disorders, na.rm = TRUE)
    )

# Creating the map 
us_map <- map_data("state") 

# Converting and fixing the data into full state names 
skin_uv_summary <- skin_uv_summary %>%
  mutate(state_name = tolower(state.name[match(state, state.abb)]))

# Joining the data from OSHA and UV df's regarding specific injury categories 
map_df <- us_map %>%
  left_join(skin_uv_summary, by = c("region" = "state_name"))

```

# Introduction

## Row 1 {height="50%"}

::: {.card title="Research Questions"}
- How does the distribution of total injuries vary across low, medium, and high UV categories?
- How do state-level UV intensity patterns compare with state-level skin disorder counts?
- How do higher UV levels relate to lost work time, including days away from work and job transfer/restriction days?
:::

## Row 2 {height="50%"}

::: {.card title="Topical Overview"}
Our project explores how OSHA tracks workplace injuries and illnesses and how these classifications influence worker protection and employer accountability.
OSHA collects company and worker data on workplace incidents across industries in the United States, including injuries caused by physical accidents, chemical exposure, radiation, and long-term health conditions.
By examining the OSHA dataset (https://www.osha.gov/establishment-specific-injury-and-illness-data) and a UV Exposure dataset (https://gis.cancer.gov/tools/uv-exposure/), our team wanted to better understand how UV exposure levels across the United States influence types of workplace illnesses and their relationship with how OSHA decides how liability is addressed and determined.
Our project focuses particularly on how OSHA measures and reports injury severity and how those classifications shape how workplace risk is understood.

To explore these questions, we analyze the OSHA Establishment-Specific Injury and Illness Dataset, which records workplace injury reports across thousands of establishments in the United States, and the UV Exposure dataset from the National Cancer Institute.
According to the U.S.
Department of Labor, OSHA was created to ensure safe and healthy working conditions by enforcing standards and collecting data about workplace injuries and illnesses.
These datasets provide insight into how different injury categories, such as skin cancer, chemical exposure, and radiation exposure, are documented and evaluated within OSHA's reporting system.
By visualizing these patterns, we can better understand how OSHA measures injury severity and how these measurements influence how workplace risks are recognized.

Our analysis focuses specifically on skin cancer and exposure-related illnesses, which represent a type of injury that often develops gradually over time.
Unlike immediate workplace accidents, conditions such as skin cancer may result from prolonged exposure to environmental factors like sunlight, chemicals, or radiation.
This raised the question: how does OSHA determine whether such illnesses should be classified as work-related conditions?
Because many environmental exposures can occur both inside and outside the workplace, identifying the origin of these illnesses can be complex.
Visualizing the dataset allows us to explore how OSHA categorizes these cases and whether certain industries appear more frequently in exposure-related injury reports.

Our primary target audience is U.S. workers, particularly those in industries with higher occupational hazards.
While OSHA data is comprehensive, it can be overwhelming and difficult for the average worker to interpret.
Many employees may never have the opportunity to explore the dataset in depth, so it is crucial to present information clearly and accessibly that highlights the injury risks specific to each job.
By translating OSHA data into actionable insights, workers can make more informed decisions about their safety and job.
Hence, this helps employers track liability and make better decisions about employee safety and risk management.
Policymakers may also find value in aggregated trends to inform changes to workplace regulations and rules.
The dataset we are analysing is the OSHA 2024 dataset, which is a collection of work-related injuries and illnesses; it is quite expansive (most companies with 10+ employees must report to OSHA), covering hundreds of companies and various sectors within those corporations.
This data set is particularly compelling because most people work or have worked; so analysing the health data, and therefore one of the aspects of safety at a job, is crucial.
Some information that would need to be added is more categories under skin illness; currently, all skin cancers, diseases, and chemical exposures are grouped into one column.
Additional data would also need to be provided on whether the injury could have been, or was, partially caused off the job.
For example, for skin-related injuries and illnesses, data such as family history of skin cancer, on- and off-the-job radiation exposure (amount, duration), and typical weather conditions (average UV Index) would be beneficial.

From an ethical perspective, this project also raises questions about how institutions measure and represent long-term or less visible health risks.
Many illnesses caused by chemical exposure, radiation exposure, or environmental factors may not show symptoms immediately and may only become detectable years after exposure occurs.
Because of this delayed visibility, systems that classify and track these injuries must make assumptions about causation and responsibility.
By visualizing patterns from UV exposure categories and the OSHA dataset, our project aims to highlight how these environmental factors, by geographical region, shape how workplace health risks are identified, measured, and addressed.
:::

# Data Context

## Row 1 {height="50%"}

::: {.card title="Dataset 1: OSHA Establishment-Specific Injury and Illness Data"}

The first dataset used in this project is the OSHA Establishment-Specific Injury and Illness Dataset, which is publicly available through the U.S.
Department of Labor's Occupational Safety and Health Administration (OSHA).
This data is publicly available here: https://www.osha.gov/establishment-specific-injury-and-illness-data.

We collected this dataset, which was published by the Occupational Safety and Health Administration (OSHA), a federal agency within the U.S.
Department of Labor responsible for ensuring safe and healthy working conditions for employees.
OSHA requires employers to report workplace injuries and illnesses each year as part of federal workplace safety regulations and government standards.
Such reports allow OSHA to monitor worker safety conditions across industries and identify trends in workplace hazards.

**Key Strengths:** The sample consists of injury and illness reports submitted by thousands of establishments and worker injury cases across various industries in the United States.
These records include information about workplace injuries and illnesses that occurred during the reporting year and were documented by employers under OSHA reporting requirements.
Each observation row represents a specific workplace establishment and includes column variables such as industry classification, injury counts, illness types, and severity categories.
The dataset records various types of workplace injuries, including incidents caused by chemical exposure, radiation exposure, environmental conditions, and other occupational hazards.

The dataset measures workplace injury outcomes by collecting information from OSHA's injury-tracking systems, which require employers to record workplace incidents in standardized categories.
These measurements include the number of injuries or illnesses recorded at an establishment, the classification of the injury type, and additional contextual information about the workplace.
The dataset contains tens of thousands of recorded establishments across different industries, providing a large sample that represents a wide range of occupational environments.

The data are collected annually and cover establishments across the United States, providing a national perspective on workplace injuries and illnesses.

**Advantages:** One benefit of this dataset is its large scale and standardized reporting structure, which allows researchers to compare injury patterns across industries and geographic regions.

**Limitations:** However, one limitation of the dataset is its reliance on employer self-reporting, which may introduce bias if injuries are underreported or misclassified.
Additionally, some illnesses—especially those that develop slowly over time—may be difficult to attribute directly to workplace exposure.
:::

## Row 2 {height="50%"}

::: {.card title="Dataset 2: U.S. Ultraviolet (UV) Radiation Exposure Data"}
The National Cancer Institute contains information about ultraviolet (UV) radiation exposure across U.S. states.
This dataset is publicly available from environmental monitoring and climate data repositories and provides geographic measurements of UV intensity.
UV radiation data is commonly used in public health research to understand environmental exposure risks for skin damage and skin cancer. The dataset is publicly available: https://gis.cancer.gov/tools/uv-exposure/.

This dataset was compiled by environmental monitoring organizations that collect atmospheric and solar radiation data using satellite observations and ground monitoring stations.

These organizations measure ultraviolet radiation levels across geographic regions to track environmental exposure patterns and potential health risks related to sun exposure.

**Key Strengths:** The sample includes UV radiation measurements recorded across multiple U.S. states and territories.
Each observation represents a geographic location and includes variables such as latitude, longitude, and average UV intensity levels.
These measurements allow researchers to compare environmental exposure levels across regions and identify areas where UV radiation may be more intense.
The dataset provides a geographic perspective for exploring potential environmental risk factors associated with skin-related health outcomes.

UV radiation levels are measured using atmospheric monitoring instruments and satellite observations that estimate UV intensity.
The dataset consists of multiple geographic records and demonstrates UV exposure patterns across the U.S.
These measurements help our team track how environmental exposures influence long-term health risks, such as skin cancer.

**Advantages:** One advantage of this data set was that it provided environmental measurements across geographic locations that could be compared when combining workplace injury data and environmental conditions to map occupational exposure risks.

**Limitations:** A limitation is that UV measurements reflect regional environmental exposure rather than individual worker exposure, so the merged data set cannot directly determine which specific illnesses were aggravated by workplace conditions.

**Data Merging and Wrangling Both the OSHA Data set and UV Exposure Data**

- We combined the OSHA injury dataset with environmental UV exposure data to explore how workplace injury reports related to skin conditions may be associated with environmental exposure patterns across regions.
- Our OSHA dataset provides information about workplace injury classification and reporting, while the UV dataset provides environmental context about exposure levels that may contribute to long-term health risks.
- Overall, both datasets allow us to examine how occupational health reporting systems interact with environmental factors that influence worker safety and management of injuries.
:::

# Data Joining and Wrangling

## Row 1 {height="50%"}

```{r}
#| title: PART 1 LOAD

library(tidyverse) 
library(readxl) #loads excel file
library(janitor) #pacakage to help clean data columns
library(ggplot2)


#loading osha_df 
osha_url <- "https://docs.google.com/spreadsheets/d/1GM90yTIXVE7xZoUrLwrqAO8yVWJsZtjNnkGbvDZhGpQ/export?format=csv&gid=1573088179"
osha_df <- read.csv(osha_url)

## Checking to make OSHA state col values in uppercase
osha_df <- osha_df %>%
  mutate(state = toupper(state))

#Loading NIH Exposure uv_df 
uv_df <- read_excel("uv-county.xlsx") %>%
    clean_names()

# Converting "statename" col in uv_df to abbreviations and matching with osha_df
uv_df <- uv_df %>%
  mutate(state = state.abb[match(statename, state.name)])

# Grouping by State-Level UV Exposure 
uv_state <- uv_df %>%
  group_by(state) %>%
  summarize(mean_uv = mean(uv_wh_m2_1961_1990, na.rm = TRUE))

## NOTE: "uv_wh_m2_1961_1990" refers to the average UV radiation measured 
## in watts-hours per square meter (WH/m^2) across years 1961-1990

#Joining both datasets
joined_df <- osha_df %>%
  left_join(uv_state, by = "state")

glimpse(joined_df)

summary(joined_df$mean_uv) 
```

## Row 2 {height="50%"}

```{r}
#| title: PART 2 ORGANIZE DATA

# Created a UV mean category in joined_df that classifies UV intensity in the state by Low, Medium or High 

joined_df <- joined_df %>%
    mutate(uv_category = case_when(
        mean_uv < quantile(mean_uv, 0.33, na.rm = TRUE) ~ "Low", 
        mean_uv < quantile(mean_uv, 0.66, na.rm = TRUE) ~ "Medium", 
        TRUE ~ "High"
    ))

joined_df

#Created a uv_summary variable that summarizes the dataframe

uv_summary <- joined_df %>% 
    group_by(state) %>% 
    summarize( 
        avg_uv = mean(mean_uv, na.rm = TRUE), 
        total_cases = n()
    )

uv_summary

# Created total skin disorders per state and avg uv variable

skin_uv_summary <- joined_df |>
    group_by(state) |>
    summarise(
        avg_uv = mean(mean_uv, na.rm = TRUE),
        skin_dis_count = sum(total_skin_disorders, na.rm = TRUE)
        )

skin_uv_summary

# Made a visualization showing the UV intensity and Skin disorder count

ggplot(skin_uv_summary, aes(x = skin_dis_count, y = avg_uv, color = factor(state))) +
       geom_point() +
       labs(
           title = "Skin Disorders & Average UV per State", 
           x = "Skin disorder count", 
           y = "Average UV", 
           color =  "State") 

# Made a visualization showing the UV category intensity and total injuries count

ggplot(joined_df, aes(x = uv_category, y = total_injuries, fill = uv_category)) +
  geom_boxplot() +
  labs(
    title = "Skin Disorders by UV Category",
    x = "UV Category",
    y = "Total Skin Disorders"
  ) +
  theme_minimal() 
```

# Visualization 

## Row 1 {height="33%"}

### Row 1 Column 1

```{r}
#| echo: false
#| title: Q1 - How does the distribution of total injuries vary across low, medium, and high UV categories?

# Creating colored violin plot 
injury_uv <- ggplot(joined_df, aes(x = uv_category, y = total_injuries, fill = uv_category)) + 
  geom_violin(color = "black", alpha = 0.7) +
  scale_fill_manual(values = c("Low" = "yellow", "Medium" = "orange", "High" = "red")) +
  labs(
      title = "Concentration of Injuries per UV Index",
      x = "UV categories",
      y = "Total injury amount"
      ) +
  theme_minimal()
ggplotly(injury_uv)

# Coding Notes:
#We used geom_violin because it shows kernal density concentration well for the purpose of analyzing injury concentration per UV category and pointing out outliers. 
#I found and learn about this function from r-charts.com.
```

### Row 1 Column 2 

```{r}
#| echo: false
#| title: Q3 - How do higher UV levels relate to lost work time, including days away from work and job transfer/restriction days?
library(ggplot2)

work_uv <- ggplot(joined_df, aes(x = mean_uv, y = total_dafw_days, color = uv_category)) +
  geom_point() +
  labs(
    title = "UV Intensity vs. Days Away From Work",
    x = "Average UV (Wh/m^2)",
    y = "Total Days Away From Work",
    color = "UV Level"
  ) +
  theme_minimal() 
ggplotly(work_uv)

days_uv <- ggplot(joined_df, aes(x = uv_category, y = total_djtr_days, fill = uv_category)) +
  geom_boxplot() +
  labs(
    title = "Job Transfer Days by UV Category",
    x = "UV Intensity Category",
    y = "Days of Job Transfer/Restriction",
    fill = "UV Level"
  ) +
  theme_minimal()
ggplotly(days_uv)

# Coding Notes: 
#I used the geom_point to create a scatterplot to visualize the relationship between two continuous variables.
```

## Row 2 {height="33%"}

```{r}
#| echo: false
#| title: Q2 - How do state-level UV intensity patterns compare with state-level skin disorder counts?


library(maps) 
library(ggplot2)
library(dplyr)

# Creating the map 
us_map <- map_data("state") 

# Converting and fixing the data into full state names 
skin_uv_summary <- skin_uv_summary %>%
  mutate(state_name = tolower(state.name[match(state, state.abb)]))

# Joining the data from OSHA and UV df's regarding specific injury categories 
map_df <- us_map %>%
  left_join(skin_uv_summary, by = c("region" = "state_name"))

# Map #1: Average UV Intensity By State
map_int <- ggplot(map_df, aes(long, lat, group = group, fill = avg_uv)) + 
    geom_polygon(color = "white", linewidth = 0.2) + 
    coord_fixed(1.3) + 
    scale_fill_viridis_c() + 
    labs(
        title = "UV Intensity Distribution Across U.S. States", 
        fill = "Average UV"
    ) + 
    theme_minimal() 
ggplotly(map_int)

# Map #2: Total Reported Number of Skin Disorder Cases By State  
map_dis <- ggplot(map_df, aes(long, lat, group = group, fill = skin_dis_count)) + 
    geom_polygon(color = "white", linewidth = 0.2) + 
    coord_fixed(1.3) + 
    scale_fill_viridis_c() + 
    labs(
        title = "Total Skin Disorders by Each State", 
        fill = "Total Skin Disorders"
        ) + 
    theme_minimal()
ggplotly(map_dis)

# Coding Notes: 
#We used the maps package, the `geom_polygon()` function from ggplot2, mutate, summarize, group_by, and left_join tools in order to analyze the geographical patterns, summarizing the UV intensity distribution across U.S. states and the total skin disorders reported by each state in the joined dataset. This helps visualize the spatial data for which regional patterns have the highest counts for UV radiation exposure. I learned how to construct the code to build these maps by reviewing examples from the UW's Technical Foundations of Informatics and other class resources.
```

# Key Takeaways and Conclusion

## Row 1 {height="33%"}

::: {.card title="Insight 1: Outliers in High-UV Regions"}

Workplace injury distributions vary across UV exposure categories.

The violin plot helped create intensity categories to map injury counts across low, medium, and high UV exposure levels.
The visualization showed that the majority of companies reported low injury counts across all UV exposure levels.
However, within the higher UV exposure categories, several noticeable outliers emerged.

This pattern suggests that while most workplaces maintain low reported injury rates, certain environments with greater UV exposure may be associated with a higher risk of overall injuries.
The general clustering of low-injury countries indicates that high reported injury frequency is uncommon among the recorded corporations.
:::

## Row 2 {height="33%"}

::: {.card title="Insight 2: Geographical Variations by State"}

In our comparison of geographical locations across the US, which reported UV intensities and skin disorder cases by state, the map visualizations showed regional patterns, with southern states having higher UV exposure levels.

Meanwhile, northern states were shown to have relatively lower UV exposure levels.
However, the reported mapping of skin disorders across US states shows that while some high-UV states report notable numbers of skin disorders, the distribution varies across states.

The map results suggest that other factors besides environmental exposure conditions influence how injuries, such as skin-related occupational illnesses, appear in OSHA's records and datasets.
:::

## Row 3 {height="33%"}

::: {.card title="Insight 3: UV Intensity Levels Have a Weak Correlation to Lost Work Time"}

Moreover, we found no strong correlation between lost work time and higher UV levels in our scatterplot.
In the boxplot, a similar pattern emerged: higher UV categories showed greater variability in lost work time, but the overall distribution was highly skewed, with a larger number of low values.
Therefore, environmental conditions and exposure do not strongly correlate with worker productivity impacts.
:::

::: {.card title="Conclusion Summary"}

**Summary**


**Limitations:** 
The data sets are aggregated at the state level which may not account for different regional variations within a state. Moreover, environmental exposure factors do not have a strong correlation in uncovering workplace safety regulations and injuries caused liabilities resulting directly from company environments. More research can be explored to assess the relationship between UV exposure and occupational health outcomes, by considering both regional variations and other categorical variables considering non-environmental factors.

**Future Research:** 
Future research can incorporate other occupational and environmental data categories, such as regional UV index data trends, outdoor workplace exposure measurements, and other worker demographics, to better visualize how environmental factors directly correlate with workplace health risks.

By combining more data categories, we can better understand how environmental exposure affects occupational illnesses and OSHA regulations.
:::

# Group Information

## Row 1 {height="50%"}

::: {.card title="Team Members"}
- **Nithi Byreddy** (The Information School, University of WA; nithib@uw.edu) is interested in information systems that design sustainable and ethical data tools and AI infrastructure.
- **Wenhao Wang** (The Information School, University of Washington; whwang1@uw.edu) is obsessed with data analysis.
- **Aryna Kaptselava** (Foster School of Business, University of Washington; akaps713@uw.edu) is interested in how government-sponsored corporations record and categorize data, and how that impacts communities.
:::

## Row 2 {height="50%"}

::: {.card title="Project Summary & Additional Information"}
**TA:** Alissa Lau

**Group Topic Statement:**

Our project helps visualize how OSHA employee-related work injuries are related to environmental UV exposure in different geographical regions and how this influences the likelihood of injury categories and company liability.
:::

